Embedded Scalars: {A}n Experimental Revisit

### 2014-05-28 text/paper/paper4.tex

Our previous regression-based analyses did not take the sequential
nature of the task into account. We initially hypothesized that
subjects might exit as early as possible, but the data from the
target-related controls suggests otherwise. In order to estimate this
bias and the preference for candidate readings, we use a simple
Bayesian cognitive model that computes the likelihood of answer
patterns for different latent sequential biases and reading
preferences. We use the data to estimate the posterior likelihood of
these latent parameters to gain insight into the likely preference
relations and the salience of local readings.

\paragraph{The model.} We outline a simple probabilistic model that
quantifies the positional bias and the competing strength of available
readings. Take the AS-conditions as an example. There are three
potential readings. The relative strength between these is given by a
probability vector $\myvec{p}^\text{as-n} =
\tuple{p^\text{as-n}_{\text{lit}},p^\text{as-n}_{\text{loc}},p^\text{as-n}_{\text{glb}}}$. We
think of $p^\text{as-n}_{\text{lit}}$, for instance, as the salience
of the literal reading for \as-sentences with neutral prosody,
relative to the global and local reading. If there were no potential
effects of sequentiality and no errors (or other readings that we
classify as errors), the vector $\myvec{p}^\text{as-n}$ would be our
prediction of observed frequencies of answer patterns. We look at four
such three-placed vectors, one for each target condition.

For the preference-related control conditions we look at three vectors
$p^k = \tuple{p^{\text{k}}_{\text{\lc}}, p^{\text{k}}_{\text{\ec}}}$,
where $k \in \set{text{ntr}, \text{\lc-cue}, \text{\ec-cue}}$. These
vectors give the relative salience of \lc- and \ec-readings for
different types of intonation. The salience of readings is independent
of the pictorial sequence. If there is a tendency to judge sentences
earlier in the sequence, rather than later, that is to be attributed
to the sequential bias, not a difference in general, task-independent
salience of readings.

To add potential effects of sequentiality, we assume that there is a
constant bias factor $q \in [0;1]$. This bias is held constant over
all conditions, because we assume that there is a global tendency to
answer early or late, independent of the sentence subjects are
confronted with. The bias factor captures whether there is a tendency
to answer early ($q > .5$) or later ($q <.5$) in the sequence. In the
\as-sequences, the literal reading can be judged first, then the local
and then the global reading. So we will assume that $p_{\text{lit}}$
is multiplied by $q$, that $p_{\text{loc}}$ is multiplied by $(1-q)q$
and that $p_{\text{glb}}$ is multiplied by $(1-q)^2$.

Finally, we also allow for the chance of subjects making mistakes. To
keep things as simple as possible (no consideration for spill-overs,
late decisions, preferences to unravel completely before judging
etc.), we assume that there is a fixed error rate $e^k \in [0;1]$ for
each sentence-intonation pair $k$. We allow for different error
rates for different sentence-intonation pairs, because we want to
allow for the possibility that non-monotonic quantifiers might be
harder to process, or that some kind of intonation (like the \ec-cue)
is harder than others. A decision can be wrong but incidentally
coincide with a reading that is indicative of a relevant
reading. Error rates multiply in proportion to the number of steps in
a sequence: more choice, more chance to be wrong.

\bigskip

HERE

Taking this together we calculate, for each critical
sentence-intonation pair $k$ a target probability vector $\myvec{t}^k
=
\tuple{t^k_{\text{lit}},t^k_{\text{loc}},t^k_{\text{glb}},t^k_{\text{err}}}$
with which we expect an answer to be classified as LIT, LOC, GLB or as
an error. For example, for AS-N, where LIT can be judged before GLB
before LOC, we get:
\begin{align*}
  t^{\text{as-n}}_{\text{lit}} & \propto q \
  p^\text{as-n}_{\text{lit}} + e^\text{as-n} &   t^{\text{as-n}}_{\text{loc}} & \propto (1-q)^2 \
  p^\text{as-n}_{\text{loc}} + e^\text{as-n} \\
  t^{\text{as-n}}_{\text{glb}} & \propto (1-q)q \
  p^\text{as-n}_{\text{glb}} + e^\text{as-n} &   t^{\text{as-n}}_{\text{err}} &
  \propto 12 e^\text{as-n}
\end{align*}
Two remarks. The vector $\myvec{t}^k$ should be normalized eventually,
which is why $\propto$ is used instead of $=$. Also, the probability
of error $t^{\text{as-n}}_{\text{err}}$ comes from the observation
that there are 15 positions in total in an AS-sequence at which
subjects can make an erroneous choice, but we have accounted for three
of these already as adding to the count of answers indicative of
relevant readings.

Similarly, we compute a target probability $\myvec{t}^k =
\tuple{t^k_{\text{late}},t^k_{\text{early}},t^k_{\text{err}}}$ for
each preference-related control condition $k$, taking into account
which reading can be judged first and the length of the respective
sequences.

>>>>

The previous regression-based analyses did not take the sequential
nature of the task into account. We initially hypothesized that
subjects might exit as early as possible, but the data from the
target-related controls suggests otherwise. The following paragraphs
spell out a simple Bayesian cognitive model that computes the
likelihood of answer patterns for different latent sequential biases
and reading preferences. We use the data to estimate the posterior
likelihood of these latent parameters to gain insight into the likely
preference relations and the salience of local readings.

\paragraph{The model.} We outline a simple probabilistic model that
quantifies the positional bias and the competing strength of available
readings. Take the AS-conditions as an example. There are three
potential readings. The relative strength between these is given by a
probability vector $\myvec{p}^\text{as-n} =
\tuple{p^\text{as-n}_{\text{lit}},p^\text{as-n}_{\text{loc}},p^\text{as-n}_{\text{glb}}}$. We
think of $p^\text{as-n}_{\text{lit}}$, for instance, as the salience
of the literal reading for \as-sentences with neutral prosody,
relative to the global and local reading. If there were no potential
effects of sequentiality and no errors (or other readings that we
classify as errors), the vector $\myvec{p}^\text{as-n}$ would be our
prediction of observed frequencies of answer patterns. We look at four
such three-placed vectors, one for each target sentence-intonation
pair.

For the preference-related control conditions we look at three vectors
$p^k = \tuple{p^{\text{k}}_{\text{\lc}}, p^{\text{k}}_{\text{\ec}}}$,
where $k \in \set{\text{ntr}, \text{\lc-cue}, \text{\ec-cue}}$. These
vectors give the relative salience of \lc- and \ec-readings for
different types of intonation. The salience of readings is independent
of the pictorial sequence. If there is a tendency to judge sentences
earlier in the sequence, rather than later, that is to be attributed
to the sequential bias.

To add potential effects of sequentiality, we assume that there is a
bias factor $q \in [0;1]$. This bias is a global parameter held
constant over all conditions, because we think of it as a general
tendency to answer early or late in the incremental verification
task. The bias factor captures whether there is a tendency to access
readings that appear earlier ($q > .5$) or later ($q <.5$) in the
sequence. In the \as-sequences, the literal reading can be judged
first, then the local and then the global reading. So we will assume
that $p_{\text{lit}}$ is multiplied by $q$, that $p_{\text{loc}}$ is
multiplied by $(1-q)q$ and that $p_{\text{glb}}$ is multiplied by
$(1-q)^2$, when determining the eventual weights of readings.

Finally, we also allow for mistakes. To keep things as simple as
possible (no consideration for spill-overs, late decisions,
preferences to unravel completely before judging etc.), we assume that
there is a fixed error rate $e^k \in [0;1]$ for each relevant
condition $k$. We allow for different error rates for different
conditions, because we want to allow for the possibility that some
sentence types (e.g., non-monotonic quantifiers) or some kind of
intonation (e.g., the \ec-cues) might be harder to process. A decision
can be wrong but incidentally coincide with an answer that is
indicative of a relevant reading. Error rates multiply in proportion
to the number of steps in a sequence: more choice, more chance to be
wrong.

Taking this together we calculate, for each pair $k$ of target
sentence type and intonation type, a probability vector $\myvec{t}^k =
\tuple{t^k_{\text{lit}},t^k_{\text{loc}},t^k_{\text{glb}},t^k_{\text{err}}}$
with which we expect an answer to be classified as literal, local,
global or as error. For example, for \as-sentences with neutral
intonation, where literal can be judged before global before local, we
get:
\begin{align*}
  t^{\text{as-n}}_{\text{lit}} & \propto q \
  p^\text{as-n}_{\text{lit}} + e^\text{as-n} &   t^{\text{as-n}}_{\text{loc}} & \propto (1-q)^2 \
  p^\text{as-n}_{\text{loc}} + e^\text{as-n} \\
  t^{\text{as-n}}_{\text{glb}} & \propto (1-q)q \
  p^\text{as-n}_{\text{glb}} + e^\text{as-n} &   t^{\text{as-n}}_{\text{err}} &
  \propto 12 e^\text{as-n}
\end{align*}
Two remarks. The vector $\myvec{t}^k$ should be normalized eventually,
which is why `$\propto$' is used above instead of `$=$.' Also, the
probability of error $t^{\text{as-n}}_{\text{err}}$ comes from the
observation that there are 15 positions in total in an \as-sequence at
which subjects can make an erroneous choice, but we have accounted for
three of these already as adding to the count of answers indicative of
relevant readings.

Similarly, we compute a target probability $\myvec{t}^k =
\tuple{t^k_{\text{late}},t^k_{\text{early}},t^k_{\text{err}}}$ for
each preference-related control sentence under intonation $k$, taking
into account which reading can be judged first and the length of the
respective sequences.
